---
title: "DRAFT -- TECHNICAL MEMORANDUM"
subtitle: "<b>TO: </b> Christine Nordahl, Olivia Surgent, Andy Dakopolos <br> <b> FROM: </b> Matt Ponzini, Sandy Taylor <br> <b> SUBJECT:</b> APP/GAIN Multivariate Longitudinal Clustering <br> <b>DATE:</b> `r paste0(' ', Sys.Date())`"
output: 
  bookdown::html_document2:
    number_sections: TRUE
    toc: TRUE
    toc_float: TRUE
    toc_depth: 3
---

<style type="text/css">

h1.title {
  font-size: 24px;
  font-weight: bolder;
}

h1 { /* Header 1 */
  font-size: 16px;
  font-weight: bold;
}

h2 { /* Header 2 */
  font-size: 14px;
  font-weight: bold;
}

h3 { /* Header 3 */
  font-size: 13px;
  font-weight: bold;
}

h3.subtitle { /* Header 3 Subtitle */
  font-size: 20px;
  font-weight: normal;
  line-height: 150%;
}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8,
                      echo=FALSE, warning=FALSE, message=FALSE)

# load packages
pacman::p_load(kml3d, ggplot2)

# update 'asis' chunk to allow inline code
knitr::knit_engines$set(asis = function(options) {
  if (options$echo && options$eval) knitr::knit_child(text = options$code)
})
  

internal <- FALSE
```

<hr style="height:1px;color:black;background-color:black">

```{r sysinfo, include=internal}
# displays R version, platform, base packages, attached packages, and packages
# loaded via a namespace
sessionInfo()
```

# STUDY DESCRIPTION/OBJECTIVE

Explore the use of Bayesian Consensus Clustering multivariate longitudinal clustering of brain region volumes in APP/GAIN participants.


# DATA PROCESSING

```{r data-import-processing, include=internal}
# if data import file(s) is (are) large, consider caching the data
#   see https://bookdown.org/yihui/rmarkdown-cookbook/cache.html for more 
#   information

source("../R/data_prep.R")
```

-   Composite volumes were created for each region by adding the two hemispheres together.  
-   Brain regions were centered and scaled.  
-   Data were reduced to complete cases.

<b>Note:</b> K-means requires complete cases for a record to be used in the algorithm. Missing values in brain region volumes will need to be imputed if we want our sample size to increase from `r nrow(brain_vol_wide)` back to `r length(unique(brain_vol_raw$subj_id))`.

# STATISTICAL ANALYSIS

Explore the longitudinal clustering of brain region volumes using BCC clustering (`BCClong::BCC.multi`). Analysis was conducted using sub-region, as well as composite region, volumes.

<b>Issues:</b>
<ul>
  <li>The model runs into a singular matrix at k=7.</li>
  <li>Can run with random initial cluster assignment, but receive an error with `mixAK` initial membership.</li>
</ul>

```{r data-analysis, include=internal}
# if analysis is computationally intensive there are a few options.
#   1) run analysis and save all results to load here
#   2) run analysis here and set chunk option cache=TRUE. A cached chunk will
#      automatically load output and objects from a previous run.
#         see https://bookdown.org/yihui/rmarkdown-cookbook/cache.html for more 
#         information
bcc_k <- readRDS("./data/bcc_models.rds")

alpha <- lapply(bcc_k, function(x) x$alpha.adjust) |> unlist()
k <- which.max(alpha) + 1

bcc_final <- BCClong::BCC.multi(
    mydat = list(
      brain_vol$Frontal_scaled,
      brain_vol$Parietal_scaled,
      brain_vol$Temporal_scaled,
      brain_vol$Occipital_scaled
    ),
    dist = c("gaussian"),
    id = list(
      brain_vol$subj_id_numeric
    ),
    time = list(
      brain_vol$scan_age
    ),
    formula = list(
      y ~ time + (1|id)
    ),
    num.cluster = k,
    initial.cluster.membership = "random",
    center = 0,
    burn.in = 2000,
    thin = 10,
    per = 1000,
    max.iter = 12000
  )
```

