---
output: 
  bookdown::html_document2:
    toc: FALSE
    number_section: FALSE
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

pacman::p_load(dplyr, tibble, tidyr, ggplot2, BCClong, cowplot)

brain_vol <- readRDS("../data-ext/brain_vol_long_20240919.rds")
bcc_model <- readRDS(file = "../data-ext/bcc_models_k2-4.rds")
bcc_hemi_model <- readRDS(file = "../data-ext/bcc_models_k2-4_hemispheres.rds")
```

# Multivariate Longitudinal Clustering of Brain Volumes using Bayesian Consensus Clustering  

Exploration of multivariate longitudinal clustering of brain region volumes using Bayesian Consensus Clustering (BCC). Analysis is conducted using the {BCClong} package (version `r packageVersion('BCClong')`).  

## Background  

Unlike other methods which provide an overall clustering that utilizes information from all features (known as global clustering), 
the BCC model provides both feature-specific clustering and global clustering. 
Feature-specific clustering is also known as local clustering as it is based on each feature alone. 
Global clustering is also known as consensus 
clustering as it is obtained by integrating information from all features to arrive at a consensus result.  

The BCC model is estimated using an MCMC algorithm,43 and similar to other model-based approaches, individuals
are assigned to the cluster to which their posterior membership probability is the largest. Missing data are assumed to be
MAR. In practice, reporting both feature-specific clusterings (L~*ir*~, i = 1, …, N, r = 1, …, R) and global clustering (C~*i*~, i = 1, …, N) is useful in understanding the 
heterogeneity underlying each feature and the heterogeneity across features. It
also facilitates the understanding of the added values of the global clustering compared to feature-specific clusterings,
particularly when the interests are in associating the clusters with a health outcome of interest.

## Analysis using Combined Brain Region Volumes (L + R Hemispheres)

### Determine Number of Clusters  

```{r}
adjusted_alphas <- lapply(bcc_model, function(x) x$alpha.adjust) |> unlist()
numb.k <- which.max(adjusted_alphas) + 1
```

The BCC was fit using scaled brain region volumes summed across hemispheres over time. We evaluated 2 to 4 clusters. The final number of clusters, as determined by the mean adjusted adherence, was `r numb.k`, \@ref(fig:alpha).

```{r alpha, fig.cap="Mean adjusted adherence from BCC models using k = 2 to 4."}
data.frame(
  k = seq(2, 4, by = 1),
  `Adjusted Alpha` = adjusted_alphas,
  check.names = FALSE
) |>
  ggplot2::ggplot() +
  ggplot2::aes(x = k, y = `Adjusted Alpha`) +
  ggplot2::geom_point() +
  ggplot2::labs(x = '*k*', y = 'Mean adjusted adherence') +
  ggplot2::theme_bw() +
  ggplot2::theme(
    axis.title.x = ggtext::element_markdown()
  )
```

### Posterior Cluster Probability

```{r}
dat <- bcc_model[[1]]$dat
dat1_uq <- dat[[1]][!duplicated(dat[[1]]$id.org, fromLast = TRUE), ]
dat1_uq$post_prob <- bcc_model[[1]]$postprob

ggplot(dat1_uq, aes(x=factor(cluster.global), y=post_prob)) +
  geom_boxplot() + 
  ggtitle("BCClong") +
  xlab("Clusters") + 
  ylab("Posterior Cluster Probability") +
  ylim(c(0,1)) +
  theme_bw() +
  theme(legend.position = "none",
        plot.title = element_text(size = 15, face = "bold"),
        axis.text=element_text(size=15),
        axis.title=element_text(size=15),
        axis.text.x = element_text(angle = 0 ),
        strip.text.x = element_text(size = 15, angle = 0),
        strip.text.y = element_text(size = 15,face="bold")
        )
```


```{r}
dat.cluster <- data.frame(
  id=dat1_uq$id.org,
  cluster.global=dat1_uq$cluster.global)

per <- round(
  100*table(dat.cluster$cluster.global)/length(dat.cluster$cluster.global),
  1
)
dat.cluster$cluster.global <- cluster.BCC <- factor(
  dat.cluster$cluster.global,
  labels =paste("Cluster ",1:numb.k," (",per,"%)",sep="")
)

dnew <- merge(brain_vol,dat.cluster, by.x='subj_id_numeric', by.y="id")
```

```{r, warning=FALSE, message=FALSE}
p1 <- ggplot(data = dnew) +
  aes(
    x = scan_age,
    y = Frontal_scaled,
    color = cluster.global,
    linetype = cluster.global,
    fill = cluster.global
  ) +
  geom_smooth(method = 'loess', se = TRUE, span = 2) +
  theme_bw()
p2 <- ggplot(data = dnew) +
  aes(
    x = scan_age,
    y = Parietal_scaled,
    color = cluster.global,
    linetype = cluster.global,
    fill = cluster.global
  ) +
  geom_smooth(method = 'loess', se = TRUE, span = 2) +
  theme_bw()
p3 <- ggplot(data = dnew) +
  aes(
    x = scan_age,
    y = Temporal_scaled,
    color = cluster.global,
    linetype = cluster.global,
    fill = cluster.global
  ) +
  geom_smooth(method = 'loess', se = TRUE, span = 2) +
  theme_bw()
p4 <- ggplot(data = dnew) +
  aes(
    x = scan_age,
    y = Occipital_scaled,
    color = cluster.global,
    linetype = cluster.global,
    fill = cluster.global
  ) +
  geom_smooth(method = 'loess', se = TRUE, span = 2) +
  theme_bw()

p_legend <- cowplot::get_legend(p1 + theme(legend.direction = 'horizontal'))

cowplot::plot_grid(
  cowplot::plot_grid(
    p1 + theme(legend.position = 'none'), p2 + theme(legend.position = 'none'), 
    p3 + theme(legend.position = 'none'), p4 + theme(legend.position = 'none'),
    nrow = 2, ncol = 2
  ),
  p_legend,
  nrow = 2,
  ncol = 1,
  rel_heights = c(0.8, 0.1)
)
```


```{r}
dnew |>
  dplyr::filter(visit == 1) |>
  dplyr::select(sex, app_diagnosis, cluster.global) |>
  gtsummary::tbl_summary(by = cluster.global) |>
  gtsummary::add_overall()

```



## Analysis using Brain Region Volumes

### Determine Number of Clusters  

```{r}
adjusted_alphas <- lapply(bcc_hemi_model, function(x) x$alpha.adjust) |> unlist()
numb.k <- which.max(adjusted_alphas) + 1
```

The BCC was fit using scaled brain region volumes summed across hemispheres over time. We evaluated 2 to 4 clusters. The final number of clusters, as determined by the mean adjusted adherence, was `r numb.k`, \@ref(fig:alpha).

```{r alpha2, fig.cap="Mean adjusted adherence from BCC models using k = 2 to 4."}
data.frame(
  k = seq(2, 4, by = 1),
  `Adjusted Alpha` = adjusted_alphas,
  check.names = FALSE
) |>
  ggplot2::ggplot() +
  ggplot2::aes(x = k, y = `Adjusted Alpha`) +
  ggplot2::geom_point() +
  ggplot2::labs(x = '*k*', y = 'Mean adjusted adherence') +
  ggplot2::theme_bw() +
  ggplot2::theme(
    axis.title.x = ggtext::element_markdown()
  )
```

### Posterior Cluster Probability

```{r}
dat <- bcc_hemi_model[[1]]$dat
dat1_uq <- dat[[1]][!duplicated(dat[[1]]$id.org, fromLast = TRUE), ]
dat1_uq$post_prob <- bcc_model[[1]]$postprob

ggplot(dat1_uq, aes(x=factor(cluster.global), y=post_prob)) +
  geom_boxplot() + 
  ggtitle("BCClong") +
  xlab("Clusters") + 
  ylab("Posterior Cluster Probability") +
  ylim(c(0,1)) +
  theme_bw() +
  theme(legend.position = "none",
        plot.title = element_text(size = 15, face = "bold"),
        axis.text=element_text(size=15),
        axis.title=element_text(size=15),
        axis.text.x = element_text(angle = 0 ),
        strip.text.x = element_text(size = 15, angle = 0),
        strip.text.y = element_text(size = 15,face="bold")
        )
```


```{r}
dat.cluster <- data.frame(
  id=dat1_uq$id.org,
  cluster.global=dat1_uq$cluster.global)

per <- round(
  100*table(dat.cluster$cluster.global)/length(dat.cluster$cluster.global),
  1
)
dat.cluster$cluster.global <- cluster.BCC <- factor(
  dat.cluster$cluster.global,
  labels =paste("Cluster ",1:numb.k," (",per,"%)",sep="")
)

dnew <- merge(brain_vol,dat.cluster, by.x='subj_id_numeric', by.y="id")
```

```{r, warning=FALSE, message=FALSE}
p1 <- ggplot(data = dnew) +
  aes(
    x = scan_age,
    y = Type2.L3.Frontal_L_scaled,
    color = cluster.global,
    linetype = cluster.global,
    fill = cluster.global
  ) +
  geom_smooth(method = 'loess', span = 2) +
  theme_bw()
p2 <- ggplot(data = dnew) +
  aes(
    x = scan_age,
    y = Type2.L3.Parietal_L_scaled,
    color = cluster.global,
    linetype = cluster.global,
    fill = cluster.global
  ) +
  geom_smooth(method = 'loess', span = 2) +
  theme_bw()
p3 <- ggplot(data = dnew) +
  aes(
    x = scan_age,
    y = Type2.L3.Temporal_L_scaled,
    color = cluster.global,
    linetype = cluster.global,
    fill = cluster.global
  ) +
  geom_smooth(method = 'loess', span = 2) +
  theme_bw()
p4 <- ggplot(data = dnew) +
  aes(
    x = scan_age,
    y = Type2.L3.Occipital_L_scaled,
    color = cluster.global,
    linetype = cluster.global,
    fill = cluster.global
  ) +
  geom_smooth(method = 'loess', span = 2) +
  theme_bw()

p_legend <- cowplot::get_legend(p1 + theme(legend.direction = 'horizontal'))

cowplot::plot_grid(
  cowplot::plot_grid(
    p1 + theme(legend.position = 'none'), p2 + theme(legend.position = 'none'), 
    p3 + theme(legend.position = 'none'), p4 + theme(legend.position = 'none'),
    nrow = 2, ncol = 2
  ),
  p_legend,
  nrow = 2,
  ncol = 1,
  rel_heights = c(0.8, 0.1)
)
```

```{r, warning=FALSE, message=FALSE}
p1 <- ggplot(data = dnew) +
  aes(
    x = scan_age,
    y = Type2.L3.Frontal_R_scaled,
    color = cluster.global,
    linetype = cluster.global,
    fill = cluster.global
  ) +
  geom_smooth(method = 'loess', se = TRUE, span = 2) +
  theme_bw()
p2 <- ggplot(data = dnew) +
  aes(
    x = scan_age,
    y = Type2.L3.Parietal_R_scaled,
    color = cluster.global,
    linetype = cluster.global,
    fill = cluster.global
  ) +
  geom_smooth(method = 'loess', se = TRUE, span = 2) +
  theme_bw()
p3 <- ggplot(data = dnew) +
  aes(
    x = scan_age,
    y = Type2.L3.Temporal_R_scaled,
    color = cluster.global,
    linetype = cluster.global,
    fill = cluster.global
  ) +
  geom_smooth(method = 'loess', se = TRUE, span = 2) +
  theme_bw()
p4 <- ggplot(data = dnew) +
  aes(
    x = scan_age,
    y = Type2.L3.Occipital_R_scaled,
    color = cluster.global,
    linetype = cluster.global,
    fill = cluster.global
  ) +
  geom_smooth(method = 'loess', se = TRUE, span = 2) +
  theme_bw()

p_legend <- cowplot::get_legend(p1 + theme(legend.direction = 'horizontal'))

cowplot::plot_grid(
  cowplot::plot_grid(
    p1 + theme(legend.position = 'none'), p2 + theme(legend.position = 'none'), 
    p3 + theme(legend.position = 'none'), p4 + theme(legend.position = 'none'),
    nrow = 2, ncol = 2
  ),
  p_legend,
  nrow = 2,
  ncol = 1,
  rel_heights = c(0.8, 0.1)
)
```


```{r}
dnew |>
  dplyr::filter(visit == 1) |>
  dplyr::select(sex, app_diagnosis, cluster.global) |>
  gtsummary::tbl_summary(by = cluster.global) |>
  gtsummary::add_overall()

```

```{r}
dnew |>
  dplyr::select(sex, visit, app_diagnosis, cluster.global) |>
  gtsummary::tbl_summary(by = cluster.global) |>
  gtsummary::add_overall()
```