---
title: "DRAFT -- TECHNICAL MEMORANDUM"
subtitle: "<b>TO: </b> Christine Nordahl, Olivia Surgent, Andy Dakopolos <br> <b> FROM: </b> Matt Ponzini, Sandy Taylor <br> <b> SUBJECT:</b> APP/GAIN Multivariate Longitudinal Clustering <br> <b>DATE:</b> `r paste0(' ', Sys.Date())`"
output: 
  bookdown::html_document2:
    number_sections: TRUE
    toc: TRUE
    toc_float: TRUE
    toc_depth: 3
---

<style type="text/css">

h1.title {
  font-size: 24px;
  font-weight: bolder;
}

h1 { /* Header 1 */
  font-size: 16px;
  font-weight: bold;
}

h2 { /* Header 2 */
  font-size: 14px;
  font-weight: bold;
}

h3 { /* Header 3 */
  font-size: 13px;
  font-weight: bold;
}

h3.subtitle { /* Header 3 Subtitle */
  font-size: 20px;
  font-weight: normal;
  line-height: 150%;
}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8,
                      echo=FALSE, warning=FALSE, message=FALSE)

# load packages
pacman::p_load(kml3d, ggplot2)

# update 'asis' chunk to allow inline code
knitr::knit_engines$set(asis = function(options) {
  if (options$echo && options$eval) knitr::knit_child(text = options$code)
})
  

internal <- FALSE
```

<hr style="height:1px;color:black;background-color:black">

```{r sysinfo, include=internal}
# displays R version, platform, base packages, attached packages, and packages
# loaded via a namespace
sessionInfo()
```

# STUDY DESCRIPTION/OBJECTIVE

Explore the use of K-Means multivariate longitudinal clustering of brain region volumes in APP/GAIN participants.


# DATA PROCESSING

```{r data-import-processing, include=internal}
# if data import file(s) is (are) large, consider caching the data
#   see https://bookdown.org/yihui/rmarkdown-cookbook/cache.html for more 
#   information

# source("../R/data_prep.R")
brain_vol_raw <- readRDS("../data-ext/brain_vol_raw_20241219.rds")
brain_vol <- readRDS("../data-ext/brain_vol_long_20241219.rds")
brain_vol_wide <- readRDS("../data-ext/brain_vol_prop_wide_20241219.rds")

prop_vol_regions <- readRDS("../data-ext/prop_vol_regions.rds")
prop_vol_sub_regions <- readRDS("../data-ext/prop_vol_sub_regions.rds")

source("../R/plot_kmean_feature.R")
```

-   Composite volumes were created for each region by adding the two hemispheres together.  
-   Brain regions were centered and scaled.  
-   Data were reduced to complete cases.

<b>Note:</b> K-means requires complete cases for a record to be used in the algorithm. Missing values in brain region volumes will need to be imputed if we want our sample size to increase from `r nrow(brain_vol_wide)` back to `r length(unique(brain_vol_raw$subj_id))`.

# STATISTICAL ANALYSIS

Explore the longitudinal clustering of brain region volumes using K-means clustering (`kml3d::kml3d()`). Analysis was conducted using sub-region proportion of total brain volume.

```{r data-analysis, include=internal}
# if analysis is computationally intensive there are a few options.
#   1) run analysis and save all results to load here
#   2) run analysis here and set chunk option cache=TRUE. A cached chunk will
#      automatically load output and objects from a previous run.
#         see https://bookdown.org/yihui/rmarkdown-cookbook/cache.html for more 
#         information

# full analysis using sub-regions
# fit model
kmean_sub <- kml3d::cld3d(
  traj = brain_vol_wide,
  idAll = brain_vol_wide$subj_id,
  time = c(1, 2, 3),
  timeInData = list(
    Frontal_L = c(2, 32, 17),
    Parietal_L = c(3, 33, 18),
    Temproal_L = c(5, 35, 20),
    Occipital_L = c(4, 34, 19),
    Frontal_R = c(7, 37, 22),
    Parietal_R = c(8, 38, 23),
    Temproal_R = c(10, 40, 25),
    Occipital_R = c(9, 39, 24)
  ),
  varNames = prop_vol_sub_regions[c(1, 3, 5, 9, 2, 4, 6, 10)]
)

# check k=2 - k=5
kml3d::kml3d(kmean_sub, nbClusters = 2:5)

# extract BIC to determine # clusters to use
BIC_sub <- rbind(
  BIC_K2 = kmean_sub@c2[[1]]@criterionValues[6],
  BIC_K3 = kmean_sub@c3[[1]]@criterionValues[6],
  BIC_K4 = kmean_sub@c4[[1]]@criterionValues[6],
  BIC_K5 = kmean_sub@c5[[1]]@criterionValues[6]
)

numb_k_sub <- which.min(BIC_sub) + 1

# extract cluster using getClusters
sub_clusters_kmeans <- kml::getClusters(kmean_sub, numb_k_sub) |>
  as.numeric()

# analysis using composite regions (Frontal, Parietal, Temporal, Occipital)
# fit model
# kmean_composite <- kml3d::cld3d(
#   traj = brain_vol_wide,
#   idAll = brain_vol_wide$subj_id,
#   time = c(1, 2, 3),
#   timeInData = list(
#     Frontal = c(14, 48, 31),
#     Parietal = c(15, 49, 32),
#     Temproal = c(16, 50, 33),
#     Occipital = c(18, 52, 35)
#   ),
#   varNames = brain_regions[c(1:3, 5)]
# )
# 
# # check k=2 - k=5
# kml3d::kml3d(kmean_composite, nbClusters = 2:5)
# 
# # extract BIC to determine # clusters to use
# BIC_composite <- rbind(
#   BIC_K2 = kmean_composite@c2[[1]]@criterionValues[6],
#   BIC_K3 = kmean_composite@c3[[1]]@criterionValues[6],
#   BIC_K4 = kmean_composite@c4[[1]]@criterionValues[6],
#   BIC_K5 = kmean_composite@c5[[1]]@criterionValues[6]
# )
# 
# numb_k_composite <- which.min(BIC_composite) + 1
# 
# # extract cluster using getClusters
# composite_clusters_kmeans <- kml::getClusters(kmean_composite, numb_k_composite) |>
#   as.numeric()
```

```{r process-results, include=internal}
## sub-regions
# process cluster assignment and merge with data
perc_sub <- paste0(
  round(
    100 * table(sub_clusters_kmeans) / length(unique(brain_vol_wide$subj_id)),
    1
  ),
  "%"
)

sub_clusters_kmeans <- factor(
  sub_clusters_kmeans,
  labels = paste0("Cluster ", 1:numb_k_sub, " (", perc_sub, ")")
)


dat_sub <- data.frame(
  subj_id = brain_vol_wide$subj_id,
  cluster = sub_clusters_kmeans
)

brain_clusters_sub <- merge(
  brain_vol,
  dat_sub,
  by = "subj_id",
  sort = FALSE
)
## composite regions
# process cluster assignment and merge with data
# perc_composite <- paste0(
#   round(
#     100 * table(composite_clusters_kmeans) / length(unique(brain_vol_wide$subj_id)),
#     1
#   ),
#   "%"
# )
# 
# composite_clusters_kmeans <- factor(
#   composite_clusters_kmeans,
#   labels = paste0("Cluster ", 1:numb_k_composite, " (", perc_composite, ")")
# )
# 
# 
# dat_composite <- data.frame(
#   subj_id = brain_vol_wide$subj_id,
#   cluster = composite_clusters_kmeans
# )
# 
# brain_clusters_composite <- merge(
#   brain_vol,
#   dat_composite,
#   by = "subj_id",
#   sort = FALSE
# )
```

# RESULTS

K-means clustering was applied to the APP/GAIN brain region data using sub-regions converted to proportion of total brain volume. We assessed the performance of 2-5 clusters and selected the final number of clusters using BIC. BIC indicated that K=2 was optimal, Figure \@ref(fig:bic).

```{r bic, fig.cap="BIC by number of Clusters"}
data.frame(
  ID = rep(1, each = 4),
  Clusters = c(rep(2:5, 1)),
  BIC = c(BIC_sub),
  Analysis = rep(c("Sub-region"), each = 4)
) |>
  ggplot2::ggplot(
    ggplot2::aes(
      x = as.factor(Clusters),
      y = BIC,
      group = ID
    )
  ) +
  ggplot2::geom_point() +
  ggplot2::geom_line() +
  ggplot2::labs(x = "Clusters", y = "BIC") +
  # ggplot2::facet_wrap(~Analysis, scales = "free_y") +
  ggplot2::theme_bw()
```

## Trajectory of the Features by Cluster

### Sub-region Analysis

```{r}
p1_L.kmeans <- plot_kmean_feature(
  data = brain_clusters_sub,
  var = prop_vol_sub_regions[1]
)
p2_L.kmeans <- plot_kmean_feature(
  data = brain_clusters_sub,
  var = prop_vol_sub_regions[3]
)
p3_L.kmeans <- plot_kmean_feature(
  data = brain_clusters_sub,
  var = prop_vol_sub_regions[5]
)
p4_L.kmeans <- plot_kmean_feature(
  data = brain_clusters_sub,
  var = prop_vol_sub_regions[9]
)
kmean_L.legend <- cowplot::get_plot_component(
  p1_L.kmeans, 'guide-box-bottom', return_all = TRUE
)


cowplot::plot_grid(
  cowplot::plot_grid(
    p1_L.kmeans + theme(legend.position = "none"), 
    p2_L.kmeans + theme(legend.position = "none"), 
    p3_L.kmeans + theme(legend.position = "none"), 
    p4_L.kmeans + theme(legend.position = "none"),
    nrow = 2, ncol = 2
  ),
  kmean_L.legend,
  nrow = 2,
  rel_heights = c(1, 0.1)
)
```

```{r}
p1_R.kmeans <- plot_kmean_feature(
  data = brain_clusters_sub,
  var = prop_vol_sub_regions[2]
)
p2_R.kmeans <- plot_kmean_feature(
  data = brain_clusters_sub,
  var = prop_vol_sub_regions[4]
)
p3_R.kmeans <- plot_kmean_feature(
  data = brain_clusters_sub,
  var = prop_vol_sub_regions[6]
)
p4_R.kmeans <- plot_kmean_feature(
  data = brain_clusters_sub,
  var = prop_vol_sub_regions[10]
)
kmean_R.legend <- cowplot::get_plot_component(
  p1_R.kmeans, 'guide-box-bottom', return_all = TRUE
)


cowplot::plot_grid(
  cowplot::plot_grid(
    p1_R.kmeans + theme(legend.position = "none"), 
    p2_R.kmeans + theme(legend.position = "none"), 
    p3_R.kmeans + theme(legend.position = "none"), 
    p4_R.kmeans + theme(legend.position = "none"),
    nrow = 2, ncol = 2
  ),
  kmean_R.legend,
  nrow = 2,
  rel_heights = c(1, 0.1)
)
```

## Features by Cluster

A quick look into the characteristics of the patients in each cluster are shown below in Table \@ref(tab:cluster-demog). Cluster 1 is predominately male, while cluster 2 has a more even male/female split.

```{r}
brain_clusters_sub |>
  dplyr::select(sex, app_diagnosis, cluster) |>
  gtsummary::tbl_summary(by = cluster) |>
  gtsummary::add_overall() |>
  gtsummary::add_p() |>
  gtsummary::modify_caption(caption = "(\\#tab:cluster-demog)Summary of Sex and APP Diagnosis by Cluster")
```